HMM

### 2015-12-29 paper3.Rnw
continuing tweaks on methods

\documentclass{article}
\usepackage{natbib}
\usepackage[utf8]{inputenc} % for accented characters
%% stuff for editing
%\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout]{changes}
%% to suppress notes & comments: \usepackage[final]{changes}
\usepackage[backgroundcolor=lightgray,textsize=tiny]{todonotes}
\usepackage{setspace}
\bibliographystyle{chicago}
\title{ HMM}
\author{Michael Li and Benjamin M. Bolker }
\date{\today}

@@

Our second goal, therefore, is to explore whether
\citeauthor{kerk2015hidden}'s results might be driven at
least in part by structural problems with the HMM, i.e.  the
assumption of temporal homogeneity in behaviour.  When large data sets
are available, information-theoretic model selection methods will
typically choose complex, highly parameterized models; when there is
only one way in which models can become more complex (e.g. by
increasing the number of behavioural modes), complexity that is
present in the data but not accounted for in the model (e.g. spatial
or temporal heterogeneity) can be misidentified as other forms of
complexity.  We predict that increasing volumes of data will
increasingly lead researchers who are accustomed to fitting relatively
small models to sparse data into such traps.  We examine whether
allowing for diurnal variation in the Florida panther data leads to
selection of models with smaller numbers of behavioural modes; we also
fit models to simulated data with varying numbers of behavioural modes
and degrees of temporal heterogeneity to test our conjecture that
heterogeneity can be misidentified as behavioural complexity.

@@

Our analysis again aims for simplicity and qualitative
conclusions rather than for picking the very best
predictive model; we focus on models that treat each step as
a univariate, log-Normally distributed observation, glossing
over both the relatively subtle distinctions in shape between
the three candidate step-length distributions and the effects
of considering multivariate (step length plus turning angle)
observations. (Note that most movement analyses, including
\cite{kerk2015hidden}, are only partially multivariate, 
treating step length and turning angle at a particular time
as independent observations and neglecting possible correlations
between them.)  However, we do briefly compare all three
step-length distributions, with and without incorporating
a von Mises-distributed turning angle in the model (Figure~1).

\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent behavioural modes.
\todo{not sure about terminology here (should be consistent about
what we call the modes/latent states/whatever}
In general, BIC values decreased as the number of states
increased from 
three to six states,
\todo{should decide on rules for numbers vs numerals (3 vs three) and be consistent}
suggesting that the six-state model was
favoured statistically. On the other hand, it was hard to interpret
the six states biologically.
We follow their lead in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the 'true' number 
of underlying movement states.  Using BIC also simplifies \ldots
\todo{cite S. Richards Ecology paper?  it's easier to see if BIC
identifies the true model than to see if AIC finds the model with
the minimum K-L distance; admit that tapering effect sizes do
happen here; claim similar results with AIC, although
generally higher number of states identified}

@@

In a HMM, the joint likelihood of emissions $\textbf{Y} =$ (\boldmath{$ y_{1},...,y_{T}$} \unboldmath) and an hidden states sequence $\textbf{Z}, z_{t} \in \{1,...,n\}, t=1,...,T$, given model parameters \boldmath{$\theta$} \unboldmath and covariates $\textbf{X}_{1:T} =$ (\boldmath{$x_{1},...x_{T}$} \unboldmath), can be written as: \\

$P(\textbf{Y}_{1:T},\textbf{Z}_{1:T}\mid$\boldmath{$\theta$}\unboldmath$,\textbf{X}_{1:T})=P(z_{1}\mid $\boldmath{$x_{1}$}\unboldmath$)P($\boldmath{$y_{1}$}\unboldmath$\mid z_{1},$\boldmath{$x_{1}$}\unboldmath$) \\ \prod\limits_{k=2}^{T}P(z_{k} \mid z_{k-1},$\boldmath{$x_{k}$}\unboldmath$)P($\boldmath{$y_{k}$}\unboldmath $\mid z_{k},$\boldmath{$x_{k}$}\unboldmath$)$ \\

where we have the following three distinct components:

\begin{enumerate}
\item Initial Probability: $P(z_{1} = i \mid $\boldmath{$x_{1}$}\unboldmath$)P($\boldmath{$y_{1}$}\unboldmath $\mid z_{1},$ \boldmath{$x_{1}$}\unboldmath $)$, the probability of state $i$ at time $t=1$ where covariate \boldmath{$x_{1}$} \unboldmath and the vector of observation density \boldmath{$y_{1}$} \unboldmath condition on covariates \boldmath{$x_{1}$} \unboldmath at state $z_{1}$.

\item Transition Probability: $P(z_{k}=j \mid z_{k-1}=i,$\boldmath{$x_{k}$}\unboldmath$)$, the probability of a transition from state $i$ at time $t=k-1$ to state $j$ with covariate $z_{k}$ at time $t=k$.

\item Emission Probability: $P($\boldmath{$y_{k}$}\unboldmath $\mid z_{k},$\boldmath{$x_{k}$}\unboldmath$)$, a vector of observation density \boldmath{$y_{k}$} \unboldmath condition on covariates \boldmath{$x_{k}$} \unboldmath at state $z_{k}$ at time $t=k$.  
\end{enumerate}
The likelihood above is the likelihood of the observed sequence given a particular
hidden sequence. In order to calculate the likelihood of the observed sequence given the
HMM, we need to sum over all possible hidden sequences. This direct method is computationally inefficient, but there exist more efficient ways to computation the likelihood and
estimate parameters numerically \citep{zucchini_hidden_2009}.

@@

We modeled three types of transition complexity instead of time homogeneous transition; extension of "Switch with covariates" by \cite{morales_extracting_2004}. For $n$ possible movement states, we have a $n \times n$ matrix that defines the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$ at time $t$ as the following multinomial logistic:

$\pi_{ij} = \exp(\eta_{ij}(t))/\left(1+\sum_{j=2}^{n}\exp(\eta_{ij}(t))\right),$ for $j={2,...,n}$

$\pi_{i1} = 1 - \sum_{j=2}^{n}\pi_{ij}$

where $\eta_{ij}(t)$ can be modelled as the following: 
\begin{enumerate}
\item Multiple block transition:  The transitioning probability $\pi_{ij}$ is a function of time (hour of day), where it is assigned to one of $M$ different time blocks:

$\eta_{ij}(t) = \sum_{m=1}^{M}a_{ijm}b_{m,t}$

where $a_{ijm}$ are parameters, and $b_{m,t}$ is a binary indicator (1 to the time block at the corresponding time $t$, and 0 otherwise).

\item Quadratic functional transition with a time covariate:

$\eta_{ij}(t) = b_{ij1}+b_{ij2}(\frac{t}{24})+b_{ij3}(\frac{t}{24})^{2}$

\item Sinusoidal functional transition with a time covariate:

$\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos(\frac{2\pi t}{24})+b_{ij3}\sin(\frac{2\pi t}{24})$

\end{enumerate}

@@

Among the different types of time inhomogeneous transition HMMs, sinusoidal time dependent transition HMMs yields the lowest BIC. The multiple block transitions was a good fit for the male panthers, using a combination of three transition matrix ($b_{1}$ for $t={1,2,3,4,5,6,21,22,23,24}$, $b_{2}$ for $ t={7,8,9,10,11,12,13,14,15,16}$, and $b_{3}$ for $t={17,18,19,20}$) and it reduced the number of states by 1 but had no effect on the females. The quadratic and sinusoidal time covariate transitions was motivated by the multiple block switching pattern attempting to capture the cyclic diurnal cycle and same complex level in terms of number of parameters. We concluded that sinusoidal time dependent transition HMM was suitable for modeling these panthers.

>>>>

\documentclass{article}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage[utf8]{inputenc} % for accented characters
%% stuff for editing
%\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout]{changes}
%% to suppress notes & comments: \usepackage[final]{changes}
\usepackage[backgroundcolor=lightgray,textsize=tiny]{todonotes}
\usepackage{setspace}
\bibliographystyle{chicago}
\title{ HMM}
\author{Michael Li and Benjamin M. Bolker }
\date{\today}

@@

Our second goal, therefore, is to explore whether
\citeauthor{kerk2015hidden}'s results might be driven at
least in part by structural problems with the HMM, i.e.  the
assumption of temporal homogeneity in behaviour.  When large data sets
are available, information-theoretic model selection methods will
typically choose complex, highly parameterized models; when there is
only one way in which models can become more complex (e.g. by
increasing the number of behavioural modes), complexity that is
present in the data but not accounted for in the model (e.g. spatial
or temporal heterogeneity) can be misidentified as other forms of
complexity.  We predict that increasing volumes of data will
increasingly lead researchers who are accustomed to fitting 
small models to sparse data into such traps.  We examine whether
allowing for diurnal variation in the Florida panther data leads to
selection of models with smaller numbers of behavioural modes; we also
fit models to simulated data with varying numbers of behavioural modes
and degrees of temporal heterogeneity to test our conjecture that
heterogeneity can be misidentified as behavioural complexity.

@@

Our analysis again aims for simplicity and qualitative
conclusions rather than for picking the very best
predictive model; we focus on models that treat each step as
a univariate, log-Normally distributed observation, glossing
over both the relatively subtle distinctions in shape between
the three candidate step-length distributions and the effects
of considering multivariate (step length plus turning angle)
observations. (Note that most movement analyses, including
\cite{kerk2015hidden}, are only partially multivariate, 
treating step length and turning angle at a particular time
as independent observations while neglecting possible correlations
between them.)  However, we do briefly compare all three
step-length distributions, with and without incorporating
a von Mises-distributed turning angle in the model (Figure~1).

\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent behavioural modes.
\todo{not sure about terminology here (should be consistent about
what we call the modes/latent states/whatever}
In general, BIC values decreased as the number of states
increased from 
three to six states,
\todo{should decide on rules for numbers vs numerals (3 vs three) and be consistent}
suggesting that the six-state model was
favoured statistically; however,
the authors used 3-state models in most
of their analyses for ease of biological interpretation.

We follow \cite{kerk2015hidden}'s lead in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the 'true' number 
of underlying movement states.  Using BIC also simplifies 
evaluation of model selection procedures; 
it is more convenient to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\citep{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \citep{BurnhamAnderson1998},
but we suspect that the qualitative conclusions stated here
for BIC-optimality will carry over to analyses using AIC instead.

@@

\newcommand{\obs}{\mathbf{y}}
\newcommand{\cov}{\mathbf{x}}
\newcommand{\state}{z}

In a HMM, the joint likelihood of \emph{emissions} (i.e., direct observations) $\mathbf Y = \obs_{1},..., \obs_{T}$ and a hidden state sequence $\mathbf{Z}, \state_{t} \in \{1,...,n\}, t=1,...,T$, given model parameters {$\boldsymbol \theta$} and covariates $\mathbf{X}_{1:T} =\cov_{1},..., \cov_{T}$, can be written as:
\begin{equation}
\begin{split}
P(\mathbf{Y}_{1:T},\mathbf{Z}_{1:T}|\boldsymbol \theta,\mathbf{X}_{1:T}) & =
P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \cdot \\
&  \quad \prod\limits_{k=2}^{T}P(\state_{k} | \state_{k-1}, \cov_{k})P(\obs_{k} | \state_{k},\cov_{k})
\end{split}
\label{eq:HMMlik}
\end{equation}

The model contains three distinct components:

\begin{description}
\item[Initial probability] $P(\state_{1} = i | \cov_{1}) P(\obs_{1} | \state_{1}, \cov_{1} )$, the probability of state $i$ at time $t=1$ where the covariate is $\cov_{1}$ and the vector of observation densities $\obs_{1}$ conditions on covariates $\cov_{1}$ at state $\state_{1}$.

\item[Transition probability] $P(\state_{k}=j | \state_{k-1}=i,\cov_{k})$, the probability of a transition from state $i$ at time $t=k-1$ to state $j$ with covariate $\state_{k}$ at time $t=k$.

\item[Emission probability] $P({\boldmath y_{k}} | \state_{k},\cov_{k})$, a vector of observation density $\obs_{k}$ condition on covariates $\cov_{k}$ at state $\state_{k}$ at time $t=k$.  
\end{description}

Eq.~\ref{eq:HMMlik} gives the likelihood of the observed sequence 
given (conditional on) a particular
hidden sequence. 
In order to calculate the overall, unconditional (or marginal) 
likelihood of the 
observed sequence, we need to average over all possible hidden sequences. 
This direct method is computationally inefficient, but there exist more efficient ways to computation the likelihood and
estimate parameters numerically \citep{zucchini_hidden_2009}.

@@

We modeled three types of transition complexity instead of time homogeneous transition; extension of ``Switch with covariates" by \cite{morales_extracting_2004}. For $n$ possible movement states, we have a $n \times n$ matrix that defines the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$ at time $t$ as the following multinomial logistic:
\begin{equation}
\begin{split}
\pi_{ij} & = \exp(\eta_{ij}(t))/\left(1+\sum_{j=2}^{n}\exp(\eta_{ij}(t))\right), \textrm{ for } j={2,...,n} \\
\pi_{i1} & = 1 - \sum_{j=2}^{n}\pi_{ij}
\end{split}
\end{equation}

where $\eta_{ij}(t)$ can be modelled as the following: 
\begin{enumerate}
\item Multiple block transition:  The transition probability $\pi_{ij}$ is a function of time (hour of day), where it is assigned to one of $M$ different time blocks:
$$
\eta_{ij}(t) = \sum_{m=1}^{M}a_{ijm}b_{m,t}
$$
where $a_{ijm}$ are parameters, and $b_{m,t}$ is a binary indicator (1 to the time block at the corresponding time $t$, and 0 otherwise).  
\item Quadratic functional transition with a time covariate:
$$
\eta_{ij}(t) = b_{ij1}+b_{ij2}(\frac{t}{24})+b_{ij3}(\frac{t}{24})^{2}
$$
\item Sinusoidal functional transition with a time covariate:
$$
\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos(\frac{2\pi t}{24})+b_{ij3}\sin(\frac{2\pi t}{24})
$$
\end{enumerate}

@@

Among the different types of time inhomogeneous transition HMMs, sinusoidal time dependent transition HMMs yields the lowest BIC. The multiple block transitions was a good fit for the male panthers, using a combination of three transition matrix ($b_{1}$ for $t={1-6,21-24}$, $b_{2}$ for $ t={7-16}$, and $b_{3}$ for $t={17-20}$) and it reduced the number of states by 1 but had no effect on the females. The quadratic and sinusoidal time covariate transitions was motivated by the multiple block switching pattern attempting to capture the cyclic diurnal cycle and same complex level in terms of number of parameters. We concluded that sinusoidal time dependent transition HMM was suitable for modeling these panthers.
